---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The following code will download **activity.zip** from:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

After that, code will unzip the file and read the CSV.

```{r echo = TRUE}

  # URL to download activity.zip
  datasetUrl = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  zipName = "activity.zip"

  # If file doesn't exist, download it
  if(!file.exists(zipName)) {
    download.file(datasetUrl, destfile="./", method="wininet", mode = "wb")
  }
  
  # Upzip activity.zip
  unzip(zipName)
  
  # Read activity.csv
  data = read.csv("activity.csv")
  
  # Head of data
  head(data)
```

Next steps are to perform the data transformation of the date column.

1. Verify data type of date column:
```{r echo = TRUE}
  # Verify data type of date column:
  str(data$date)
```

2. Transfor the date column to Date data type:
```{r echo = TRUE}
  # Transform from Factor to Date
  data$date = as.Date(data$date, format = "%Y-%m-%d")
```

3. Verify data type of date column:
```{r echo = TRUE}
  # Verify data type of date column:
  str(data$date)
```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day
```{r echo = TRUE}
  # Calculate the steps by day
  stepsByDay = aggregate(steps ~ date, data, sum)

  # Head of stepsByDay
  head(stepsByDay)

  # Validate NA for stepsByDay
  any(is.na(stepsByDay))
```

2. Make a histogram of the total number of steps taken each day
```{r plot, fig.width=10, echo = TRUE}
 
  # Create plot steps by day
  plot(stepsByDay$date, stepsByDay$steps, 
       type="h", 
       main="Histogram of Steps by Date",
       xlab="Date", 
       ylab="Steps", 
       col="blue",
       lwd=5
       )

  # Mean of the steps
  abline(h=mean(stepsByDay$steps), col="red", lwd=2)
```

```{r histogram, fig.width=10, echo = TRUE}
  # Create histogram with the frequency of steps
  hist(stepsByDay$steps, main = "Total Steps by Day", col="blue", xlab="Number of Steps")
```


3. Calculate and report the mean and median of the total number of steps taken per day
```{r echo = TRUE}
  # Summary of steps by day
  summary(stepsByDay)

  # Mean of steps
  mean(stepsByDay$steps)

  # Meadian of steps
  median(stepsByDay$steps)

```


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
