---
title: "The Most Harmful Natural Events to Population Health and the Economic Consequences"
author: Ignacio Sandoval
output: 
  html_document:
    keep_md: true
---

## Synopsis

events are most harmful to population health
events have the greatest economic consequences


## Data Processing

### Requiered Libraries

Below are the libraries requiered to perform the data transformation and analysis.

* lubridate
* dplyr
* ggplot2

```{r echo=FALSE, message=FALSE}
  # Install lubridate library
  if(!require(lubridate)){
    install.packages("lubridate", repos = "https://cran.rstudio.com/")
  }

  # Install dplyr library
  if(!require(dplyr)){
    install.packages("dplyr", repos = "https://cran.rstudio.com/")
  }

  # Install ggplot2 library
  if(!require(ggplot2)){
    install.packages("ggplot2", repos = "https://cran.rstudio.com/")
  }
```

### Loading the data

The following code will download the file **StormData.csv.bz2** from:

* [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb]

After file has been downloaded, code will read the compressed CSV.

```{r echo = TRUE}

  # URL to download StormData.csv.bz2
  datasetUrl = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  zipName = "StormData.csv.bz2"
  
  # If file doesn't exist, download it
  if (!file.exists(zipName)) {
    download.file(datasetUrl,
                  destfile = zipName,
                  method = "wininet",
                  mode = "wb")
  }

  # Read StormData.csv
  data = read.csv(zipName)
  
  # Head of data
  head(data)
```

### Data transformation

**Transform date/time variables**


```{r echo = TRUE}

  # Upper case Time Zone
  data$TIME_ZONE = toupper(data$TIME_ZONE)

  # Convert date column
  data$BGN_DATE = mdy_hms(data$BGN_DATE)
  
  # Format time column
  fTime = function (d, t, tz) {
    d = toString(d)
    t = toString(t)
    tz = toString(tz)
    
    if (nchar(t) == 4) {
      if (t == "0000" | t == "2400" | t == "000") {
        dt = paste(d, "00:00:00")
      } else {
        dt = paste(paste(d, sub("([[:digit:]]{2,2})$", ":\\1", t)), ":00", sep = "")
      }
    } else {
      dt = paste(d, t)
    }
    
    return(paste(dt, tz))
  }
   
  data$BGN_DATETIME_STR = mapply(fTime, data$BGN_DATE, data$BGN_TIME, data$TIME_ZONE)
   
  data$BGN_DATETIME = ymd_hms(data$BGN_DATETIME_STR)
  
  # subset = data[is.na(data$BGN_DATETIME),]

```

**Transform Event Type**



```{r echo = TRUE}

  # Upper case Type of Event
  data$EVTYPE = toupper(data$EVTYPE)

  # Subset of FATALITIES and INJURIES by Event Type
  eventsFatal = subset(aggregate(FATALITIES ~ EVTYPE, data, sum), FATALITIES > 0)
  eventsInjur = subset(aggregate(INJURIES ~ EVTYPE, data, sum), INJURIES > 0)
  harmfulEvents = full_join(eventsFatal, eventsInjur, by = "EVTYPE")
  
  # Create main groups of events types
  key = c("AVALANC", 
          "COLD", "FREEZ", "FROST", "HYPOTHERMIA", "HYPERTHERMIA", "LOW TEMPERATURE", "ICE", "ICY", 
          "HEAT", 
          "HAIL", "GLAZE",
          "FLOOD", "FLD", "RAPIDLY RISING WATER",
          "FOG", 
          "FIRE",
          "RAIN",
          "HEAVY SURF", "HIGH SURF",
          "RIP CURRENT", 
          "MUDSLIDE", "LANDSLIDE",
          "HURRICANE", 
          "LIGHTNING", 
          "SNOW", "SLEET", "WINTRY MIX",
          "TORNADO", 
          "TSUNAMI",
          "THUNDERSTORM", "THUNDER STORM", "STORM",
          "WARM", "WIND", "WINTER")
  
  value = c("AVALANCHE", 
            "COLD", "COLD", "COLD", "COLD", "COLD", "COLD", "COLD", "COLD",
            "HEAT", 
            "HAIL", "HAIL",
            "FLOOD", "FLOOD", "FLOOD", 
            "FOG",
            "FIRE",
            "RAIN",
            "HEAVY SURF", "HIGH SURF",
            "RIP CURRENTS", 
            "SLIDES", "SLIDES",
            "HURRICANE",   
            "LIGHTNING", 
            "SNOW", "SNOW", "SNOW", 
            "TORNADO",
            "TSUNAMI",
            "THUNDERSTORM", "THUNDERSTORM", "STORM",
            "WARM", "WIND", "WINTER")
  
  mapEvents = data.frame(key, value)
  
  # Function to create a class based on text
  funClass = function(value) {
  
    map = as.data.frame(mapEvents)
    
    for (i in 1:nrow(map)) {
      v = map[i,]
      k = toString(v[,1])
      v = toString(v[,2])
      
      if(grepl(k, value)) {
        res = v
        return(res)
        break;
      }
    }
    
    return("OTHER")
    # return(value)
  }
  
  # Clean subset of FATALITIES and INJURIES by Class (Main event Type)
  harmfulEvents$CLASS = as.factor(unlist(lapply(harmfulEvents$EVTYPE, funClass)))
  eventsFatalClean = subset(aggregate(FATALITIES ~ CLASS, harmfulEvents, sum), FATALITIES > 0)
  eventsInjurClean = subset(aggregate(INJURIES ~ CLASS, harmfulEvents, sum), INJURIES > 0)
  harmfulEventsClean = full_join(eventsFatalClean, eventsInjurClean, by = "CLASS")

```



```{r echo = TRUE}
  # subset = data[245000:250000,]

  
  
  

```



## Results

### Most harmful events to population health



### Events with the greatest economic consequences





